---
title: "QM Programming Basics 2 - More variables,  plots, data files"
output:
  html_document:
    df_print: paged
---

We talked last class about data types. There are 5 main variable types (or classes)  for individual objects in R: character, numeric (real or decimal), integer, logical, complex.

Logical variables have values of TRUE or FALSE
Integers and numeric data are stored differently, but these days storage isn't such a big deal and R doesn't require users to worry about the difference - R handles the differences in the background.

There are additional classes that build on individual objects: vectors, lists, matrices, arrays, factors, data frames, date, time series, ... 
We'll be talking about many of these over the course of the semester.

As we discussed last class, there are several ways to create a vector

```{r}
x = 8:10  #using ":" - increments by 1
y = seq(12.2,14.5,0.2)  #sequences with equal increments
z = c(12.5,13.2,14.7)   #any list of items
```

There are some particular vectors that can be useful. We can create a vector of 0's or 1's (or anything else) using rep():

```{r}
a = rep(0,10)
b = rep(1,3)
print (a)
print (b)
```

Last class we tried adding 2 vectors

```{r}
s = x + z
print(s)
```

We can also multiply and divide:

```{r}
m = x * z
print(m)
```

Note that these operations work element-by-element. [There are other ways to multiply vectors.]

I was careful to select vectors of the same length. What if they are not same length?

```{r}
s = x + y
print(s)
```

Be careful! R may give a warning but will also give an answer even if it isn't what you intended.

There are many built-in functions in R that we can use to calculate more complicated variables.

Let’s make a time vector t that goes from 1 to 365 by 7

First value? t[1]. In general following vector by [] with an integer i in [i] gives the value of the vector at the ith position in the vector. i is called an index.

We can use a range of indices in the []. E.g., t[10:20]

```{r}
t = seq(1,365,7)
print(t[1])
print(t[10:20])
```

Last value? t[length(t)]. 
Not as nice as matlab (t(end)) or python (t(-1))

Every other value? i = seq(1,53,2), t[i]

```{r}
print(t[length(t)])

i=seq(1,53,2)
print(t[i])
```

We'll use an equation with t as independent variable to calculate a new (dependent) variable f: 

Let f = 15 * sin(2*pi*t/365) + 12

Note R has built-in value for pi (like matlab), 
built-in math functions like sin
Sin assumes argument is in radians! 
For sin(90 deg) use sin(90*pi/180)

```{r}
f = 15 * sin(2*pi*t/365) + 12
```

How long will f be? Same length as t.

Plot(t,f) [Note t and f have to be same length to plot]

```{r}
plot(t,f)
```

Improve axis labels:
  plot(t,f,xlab=”Time”,ylab=”Temperature, deg-C”)
  
Changle symbol type:
  plot(t,f,xlab=”Time”,ylab=”Temperature, deg-C”, pch=3)     
  [Help pch]
  
Change from symbols to line:
  add: type="l" [default type = “p” (points)]
  
Change line type:
  lty (line type, 1-6), 
  lwd (line width – default = 1), 
  col (color, with standard color names, e.g., red, blue)    
  [colors() returns full list).

Overplotting line and points: add type=”o”

Make a prettier version of the plot above. E.g.,
```{r}
plot(t, f, xlab="Year day", ylab="Temperature, deg-C", pch=3, type="o", col="red")
```

```

3.	Scripting in R

Enable script window in R Studio (upper left)
Type in commands for btplt.R

```{r}
# Calculate mean travel time
bt=read.csv(file="bt.csv",sep=",",header=TRUE)
t = bt[,1]
c = bt[,2]
tbar = sum(t*c)/sum(c)
print(tbar)
```

Save
To run, click Source button on upper right of source editor or highlight all text and click run or press Ctrl+Shift+Enter



